---
title: Grade Retention and Mixed Race
author: 
  - Aaron Gullickson, *University of Oregon*
abstract: This is a test abstract
thanks: The corresponding author can be reached at aarong@uoregon.edu. All code and data for this project are available at ...
format:
  pdf:
    documentclass: scrartcl
    papersize: letter
    fig-height: 4
    fig-pos: t
    keep-tex: false
  docx:
    toc: false
bibliography: ../project.bib
execute: 
  eval: true
  echo: false
  warning: false
  error: false
editor: source
---

```{r setup}
#| include: false
library(here)
source(here("analysis","check_packages.R"))
source(here("analysis","useful_functions.R"))
```

```{r load-data}
load(here("analysis","output","acs.RData"))
load(here("analysis","output","coef_table.RData"))
load(here("analysis","output","monoracial_results.RData"))
# change names of models for display
coef_table <- coef_table %>%
  mutate(model=factor(model,
                      levels=c("crude","fixed effects","cultural resources",
                               "parent resources"),
                      labels=c("Raw","Baseline model","Cultural model",
                               "Full model")))
```

# Introduction

The experiences of mixed race individuals will play an important role in the future of race in the US. A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines (==CITE==). This demographic phenomenon may have a potentially transformative effect on racial boundaries. Although race mixing has always been a feature of US history (and the history of any racialized social system) (==CITE==), this biracial boom is notable for its demographic size and for occurring in  post-Civil Rights ear in which historical norms govern how to handle racial mixture may be dissolving or shifting. Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into the prevailing racial hierarchy that reinforces a system of racial inequality.

Extensive scholarship documents the identification and classification decisions of multiracial individuals (==CITE==). ==SUMMARIZE BRIEFLY==. On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers. The lack of work on this topic, despite its acknowledged relevance, results from two important limitations in most available data sources. 

First, using direct identification of biracial respondents on data sources such as the Census or large scale surveys is problematic. Research on the identification and classification of biracial individuals indicates that many individuals whose parents belong to different races will not always or consistently identify with both races (==CITE==). Social status and social mobility has been shown to predict racial identification (==CITE==) and this process of reverse causation is likely to be strongest for more ambiguous groups such as biracials. Thus, differences in outcome may in some part produce differences in identification rather than the other way around when examining biracial individuals identified by self-classification.

This problem of reverse causation can be largely addressed when researchers have access to the racial identification of a respondent's biological parents. While such information is available on some large scale longitudinal surveys, researchers then run into a second problem of sample size. The samples of biracial respondents in such cases are generally far too small to make reliable statistical inferences. This limitation is particularly severe for biracial individuals with two non-white parents.

In this article, I exploit features of the American Community Survey to identify one life outcome that is available in large scale Census data - grade retention. Being held back in school is a relatively common occurrence that has consequences for later educational transitions and is disproportionately applied by race (==CITATIONS==). Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience in their life and within it we can clearly see the imprint of the racial hierarchy. Because, this outcome occurs to children living in an often natal household, I can identify the race of both biological parents for many cases. This identification allows me to compare the likelihood of grade retention for biracial children in comparison to their monoracial peers. Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents. 

The results provide a window into the placement of biracial children within an existing system of racial inequality in our educational systems. ==PITHY SUMMARY==

# Background

How might we expect the outcomes for biracial individuals do differ from their monoracial peers? In every case, the relevant comparison is how a biracial person fares relative to the two constituent groups with which their parents identify. For example, in the case of an individual with one Black and one White parent, we would compare the average outcomes for such individuals to the average outcomes for individuals with two White and two Black parents. The outcome for the Black/White group may be  similar to one of the monoracial groups, it may me somewhere in between the outcomes for the two monoracial groups, or it may be higher or lower than either of the monoracial groups. 

The most direct pathway for understanding the placement of this outcome is via how biracial individuals are classified by those in positions of power and control, such as teachers, administrators, and employers. To the extent that biracial individuals are classified as more like one of their monoracial constituent groups. they will receive the same treatment and either reap similar rewards or suffer similar discrimination and penalties. 

Two well known norms have long governed such patterns of identification. The pattern of *hypodescent* dictates that individuals with mixed racial ancestry will be classified as members of the lower status group while the pattern of *hyperdescent* dictates that such individuals will be classified as members of the higher status  (==CITATIONS==) Hypodescent, in the form of the "one drop rule" has long characterized the experiences of mixed part-Black individuals who are treated as exclusively Black (==CITE F JAMES DAVIS==). The pattern of hyperdescent blends mixed race individuals into the often unmarked higher status group and is thus more difficult to see in practice. Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming to be White (==CITATIONS==). ==NEED TO MIX IN SOME OF THE COMMENTS IN ANN AND ALIYAS RECENT PAPER==

Norms of hyperdescent and hypodescent may account for cases where the biracial group has outcomes similar to one of the monoracial constituent groups. However, two issues may diminish the relevance of these norms for understanding the placement of contemporary biracial populations. First, for individuals with Asian and Latino ancestry, no historical norm governs the placement of individuals with mixed race. Second, the historical norms for those of part-Black and part-Indigenous ancestry may also be breaking down (CITATION). MORE TO DO HERE In both of those cases, it is not entirely clear if such norms exist outside of cases in which the other component in the mixing is Whiteness. Thus, while these norms may give us some historical expectations for certain groups, they are of limited predictive value more generally.

Patterns of hyperdescent and hypodescent expects biracial individuals to have outcomes more similar to one monoracial group. Another possibility is that biracial individuals will have outcomes that are somewhere between their two constituent monoracial groups. Why might such an in-between status exist? There are several possibilities.

Racial ambiguity in physical appearance may help to produce an in-between status. This racial ambiguity may protect biracial respondents from the worst stigmatization and discrimination in social settings. However, by the same coin, it may allow them to benefit less from membership in advantaged groups. The mechanisms for this process may be two-fold. Racial ambiguity may "soften" any particalar experience of discrimination because the person in question does not conform to an ideal type and therefore may be less likely to trigger a sense of racial threat or antipathy. Additionally, racial ambiguity may contribute to more probabilistic and uncertain assessments of racial identification by various social actors that contribute to a lower sum total of disadvantage. 

Notably, both of these mechanisms could co-exist with norms of hypo/hyperdescent. We know, for example, that lighter skin tone among monoracial Black individuals is associated with better outcomes (==CITATION==). So, while the one-drop rule provides a consistent classification system, it does not create uniform experiences for all Blacks and skin tone may help to "soften" the experience of discrimination. Similarly, if an in-between status results for biracial individuals becaise they are less consistently classified, that does not mean that the classification system and the norms that underlie it are weakening. (==CITE MONK HERE==)

Aside from ambiguity in physical appearance, a variety of non-material symbolic resources (e.g. social and cultural capital) may also play a role in producing an in-between status. We know that racial disparities persist across a variety of outcomes even after controlling for a wide variety of socioeconomic variables (==CITATIONS==). The residual effect of race here is a "black box" that we often think of as compromising some combination of (a) discrimination and (b) other forms of social and cultural capital not well captured in our data that are differently distributed by race. Biracial individuals may inherit this black box of unobserved resources in weights related to their parent's races, thus helping to produce results that are in-between the two constituent groups. (==I SHOULD GIVE SOME EXAMPLES AND CITATIONS OF WHAT I AM TALKING ABOUT HERE==)

All of these arguments relate to where we might expect the outcomes of biracial individuals to fall relative to their monoracial peers. However, the experience of hybridity itself may have positive and negative effects that do not tell us the location itself but rather may modify it in the context of other patterns. On the positive side, hybridity may enlarge the cultural "toolkit" (==CITATION==) of biracial individuals. This enlarged toolkit may allow such individuals to more easily navigate a variety of social situations (e.g. code switch). On the negative side, Stonequist's (==CITATION==) "marginal man" hypothesis suggests that the dissonance of an identity that crosses strong and salient racial boundaries will be experienced by biracial individuals as not fitting in anywhere. This may result in negative mental health experiences for biracial individuals that can spill over into other types of outcomes.

==NO FORMAL HYPOTHESIS==


-   In all of this, its crucially important to understand how other resources affect our understanding of the placement of biracial individuals. Biracial individuals are the products of interracial unions and individuals are not randomly selected into such unions. This can lead to underlying differences in the socioeconomic resources that biracial individuals possess relative to their monoracial peers. However, this selection can be quite complex. (DISCUSS CHENG AND LIVELY HERE?)

    -   Status exchange suggests that members of lower status racial groups will be positively selected into interracial unions, while higher status racial groups will be negatively selected. How this affects the relative resources of biracial individuals will depend then in some part on the relative strength of these phenomena, the gender balance in a given form of interracial unions, and how the gender of parents affects outcomes. Status exchange may also not be present in all forms of interracial unions.

-   Review previous studies on outcomes

    -   Campbell
    -   Herman
    -   Holloway/Ellis stuff
    -   Udry? - also related stuff by Cheng? How does Cheng and Powell fit in here?
    -   Do some more bibliographic work, particularly forward searching to see if I am missing something big here.
    -   Limitations of this work
        -   Small sample sizes lead to inconsistent estimates in most of the survey data.
        -   In many cases, impossible to understand outcomes for part-Latino biracials.

-   Why grade retention as an outcome?

    -   Grade retention is a relatively common practice. Discuss overall trends in grade retention and connection to educational reforms.
    -   Grade retention has negative effect on later outcomes, even controlling for cognitive differences (CITE WORK HERE).
    -   Clear inequalities by race in practices of grade retention as well as by SES (CITE WORK).
    -   In effect, grade retention may be for many individuals the first encounter with the wider stratification system that sorts individuals and helps to perpetuate inequality.

-   What I do and how it improves our understanding.

# Data and Methods

Data for this analysis come from the American Community Survey (ACS), an annual 1-in-100 survey of the US population, produced by the Census Bureau. To increase sample size for smaller populations of biracial respondents, I pool ACS data for a full decade from 2010-2019. While pooling data helps to increase sample size, it may generate biases if the likelihood of grade retention is changing over time in complex ways. I address this issue in the modeling techniques discussed below. All data were extracted from the Minnesota Population Center's Integrated Public Use Microdata Series [@ruggles_ipums_2020].

Ideally, I would restrict the sample only to children living in a household with two biological parents. However, the ACS only records the direct family relationship between the head of household andother members of the family. Since 2008, the ACS also distinguishes between biological children, adopted children, and stepchildren. Respondents can therefore be identified as biological children of another member of the household only if that member is the head of household or the biological child of the head of household. Determining whether the partner/spouse of that biological parent is the other biological parent is more difficult.

Prior work has used a variety of additional restrictions to limit the analysis to children who are more likely to be the biological children of both parents [@saenz_persistence_1995; @xie_racial_1997]. Some of those procedures rely upon an assumption that most parents would be married that was more tenable in historical data than contemporary data. Therefore, I do not explicitly restrict by marital status, I do restrict cases in two other ways that are consistent with prior work. First, I restrict the sample to cases where both parents were of a reasonable age at the birth of the child (aged 15-44 for mothers and aged 15-60 for fathers). Second, I restrict the sample to only those cases where the reported race and Hispanicity of the child is consistent with at least one of the parents.

The final analytical sample consists of children aged 5-20 who are currently enrolled in the K-12 school system and live in a household with two parents who are both likely to be biological parents. This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren, limiting it exclusively to households with two parents present and over-representing the proportion of cases in which those parents are married. The result of this bias is well-known. Children living in two-parent nuclear households have greater educational success than other children as a result of material, social, and cultural resources (CITATIONS). Therefore, this sample restriction will upwardly bias estimates of grade retention. However, this bias is less problematic for the goal of this study, which is to estimate differences between groups. The sample restriction effectively eliminates differences in family structure across racial groups which may account for some observed differences in grade retention. However, the remaining differences observed here are likely to hold for the full population in the absence of strong racial differences in the effect of family structure on grade retention.

The race of children is calculated by a cross-tabulation of parent's races. A full cross-tabulation of the race/Hispanicity of each parent would produce far too many combined categories to be analytically useful. Therefore, I collapse each parent's race into the categories of White, Black, Asian, Latino, and Indigenous. These categories form the "ethnoracial pentagon" that is commonly used in popular practice and government tabulation to identify race in the US [@hollinger_postethnic_1995]. Respondents are identified as Latino base on their response to the Hispanicity question, regardless of their response to the race question. Respondents are identified as Indigenous if they responded as either American Indian/Alaska Native or as Pacific Islander. Because the goal of my analysis is to examine specifically the outcomes of "first-generation" biracial children, I exclude cases where at least one parent identified with multiple races.

The cross-tabulation of parent's races leads to ten distinct biracial categories. The total sample size for each of these biracial categories is shown in @tbl-descriptive, along with several descriptive statistics. The sample sizes are quite large for many of the groups, suggesting substantial statistical power in determining the relative outcomes for these groups. Even among the smallest group of Indigenous/Asian, the sample size is substantially greater than one would see in typical survey data.

```{r}
#| label: tbl-descriptive
#| tbl-cap: Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.
tbl_summary <- acs %>%
  group_by(race) %>%
  summarize(n=n(), 
            held_back=percent(mean(below_exp_grade), accuracy=0.01),
            mean_age=round(mean(age), 1),
            f_born=percent(mean(foreign_born),  accuracy=0.1),
            par_fb=percent(mean(foreign_born_father | 
                                  foreign_born_mother), accuracy=0.1),
            fam_inc=dollar(round(mean(family_income), 0)),
            own_home=percent(mean(own_home), accuracy=0.1),
            par_ba=percent(mean(degree_father=="BA" | 
                                  degree_father=="G" | 
                                  degree_mother=="BA" | 
                                  degree_mother=="G"), accuracy=0.1)
  ) %>%
  arrange(desc(n)) %>%
  mutate(n=comma(n))

index_biracial <- which(str_detect(tbl_summary$race, "/"))

tbl_summary %>%
  kbl(booktabs = TRUE,
      linesep = NULL,
      align = c("l",rep("r", ncol(tbl_summary)-1)),
      col.names=c("race",
                  "n",
                  "clearly below expected grade", 
                  "mean age",
                  "foreign-born",
                  "either parent foreign-born",
                  "mean family income",
                  "own home",
                  "either parent, college degree"
      )) %>%
  kable_styling(font_size=8) %>%
  column_spec(3:10, width = "1.2cm") %>%
  row_spec(index_biracial, background = "#D3D3D3")
#%>% landscape()
```

To measure grade retention, I compare the age of respondents to their current grade in school. Because the ACS lacks data on the exact birthday of respondents, it is not possible to exactly identify students who are below the expected grade level. A student of any given age in years can always be expected in two different grades. For example, a first grader is expected to be either six or seven years of age, while a second grader is expected to be either seven or eight years of age. Therefore, with information only on current grade and age in years, it is impossible to determine whether a seven year old in first grade is behind the expected grade level.

Instead of trying to identify exact grade retention, I instead identify cases in which students are clearly below expected grade. A student is clearly behind expected grade when the age of the student is greater than either of the possible ages for a student of that grade (e.g. a first grader who is eight years old). This measure will underestimate the total proportion of students that have been held back. However, the goal of this study is to examine differences across groups rather than total levels. The only factor determining this bias is whether the birthday of the student takes place earlier or later in the year, and is unlikely to vary substantially across groups.[^1]

[^1]: Biracial respondents are slightly younger than monoracial respondents in the sample as a whole because interracial unions are becoming more common over time HOW STRONG COULD THE BIAS BE HERE AND WHAT CAN I DO ABOUT IT?

==ACKNOWLEDGE POTENTIAL ISSUE WITH DROPOUT STUDENTS==

@tbl-descriptive shows the percent of each monoracial and biracial group that is clearly below the expected grade. My goal is to understand the degree of grade retention for each biracial group relative to its constituent monoracial groups (e.g. a Black/White student compared to White and Black students). Throughout this article, I use a visual approach to illustrate this placement.

As an example, I display the percent of students clearly behind grade for White, Black, and Black/White students in @fig-example. The point estimate and confidence interval for the Black/White group are shown in red, while the two monoracial groups are shown via grey dots and confidence bands. For all three cases, I use 83.4% confidence intervals, because overlap on such intervals indicates that the observed difference between the two measures in the sample is not statistically distinguishable from zero in the population at $p<.05$ (CITE).[^2] In this case, the lack of overlap between any of the confidence bands indicates that the Black/White probability of being clearly below expected grade is statistically distinguishable from both monoracial probabilities.

[^2]: The correspondence to $p<.05$ is only approximate due to differences in the two standard errors between estimates. However, this approximation will be extremely close in almost all cases.

```{r example-plot}
#| label: fig-example
#| fig-cap: Probability of being clearly below expected grade for biracial black/white respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
g <-  coef_table %>%
  filter(mrace=="Black/White" & model=="Raw") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial & 
                               mrace=="Black/White" &
                               model=="Raw"))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade")

g+annotate("text", 1, 0.035, hjust="left",
             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
  annotate("text", 2, 0.045, hjust="left",
             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
  annotate("text", 2.5, 0.036, hjust="left",
             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
```

Additionally, I include a measure of the "halfway" point between the two racial groups.[^3] This halfway point allows me to determine where the biracial group falls relative to the expectation of being halfway between the two constituent monoracial groups. In this case, Black/White students are much closer to White students in their risk of grade retention, and have probabilities much lower than both Black students and the halfway expectation.

[^3]: The standard error of this halfway estimate used to construct the confidence interval is given by $\sqrt{(s_1^2+s_2^2)/2}$ where $s_1$ and $s_2$ are the standard errors of the estimates for the two monoracial groups.

@fig-example shows the raw differences between the three racial groups of interest. However, in practice, I want to estimate differences across groups while holding constant a variety of potentially confounding variables. To do this, I estimate a set of logit models that predict the likelihood of a student being clearly behind expected grade by race and a variety of other variables. I then construct figures similar to @fig-example by calculating from the model the average predicted probabilities (APP) for each racial group. APPs estimate the average probability of the outcome by some covariate across all cases, while holding constant all other variables. APPs are akin to average marginal effects (AME) and estimated in the same way via the delta method. Whereas AMEs estimate differences or slopes, APPs estimate the level for a given category.[^4]

[^4]: Output from the full models used to construct these APPs is available in the supplementary materials provided with this article.

I begin with a baseline model that adjusts for a variety of nuisance characteristics that may vary across groups and thus need to be controlled in all models. First, I include fixed effects for state of residence because states dictate educational policy and thus can differ substantially in the likelihood of grade retention. I cannot identify specific school districts, but I also include dummy variables indicating whether the student lived in a central city, suburban, or rural area.[^5]

[^5]: The location cannot be determined for all students and I therefore also include a fourth category of unknown in all models.

The probability of being clearly below expected grade also increases with the student's current grade, so I include fixed effects for the current grade of the student. Additionally, @fig-year-grade shows that the percent of students clearly behind grade level has declined over time, but this decline has been much more substantial at higher grade levels. For the elementary grades, there is no evidence of a decline at all. To account for this grade-specific decline in the models I include an interaction between a linear year term and current grade.[^6]

[^6]: Sensitivity analysis showed that this functional form was preferred by the Bayesian information criterion to models with no interaction and a model with interaction terms that treated year as a categorical variable. I also considered similar interactions between state and year, but none of the models with these interaction terms were preferred.

```{r}
#| label: fig-year-grade
#| fig-cap: Trends over time and grade in the percent of students clearly behind grade level. Lines are fit to each set of points by grade via lowess smoothing.
  
acs %>%
  group_by(current_grade, year) %>%
  summarize(below_grade=mean(below_exp_grade)) %>%
  ggplot(aes(x=year, y=below_grade, group=current_grade, color=current_grade))+
  #geom_line()+
  geom_smooth(se=FALSE)+
  geom_point()+
  scale_x_continuous(breaks=c(2010,2014,2018))+
  scale_y_continuous(labels=scales::percent)+
  labs(x="year", y="percent clearly behind grade level for age",
       color="current grade")+
  theme_bw()
```

The baseline model also includes the race of the respondent. I then add additional terms measuring material and cultural resources that may account for racial differences in a set of subsequent models. First, I include measures of nativity and English proficiency for both the respondent and each of their parents. Second, I include a categorical measure of highest degree earned for each parent. Finally, I include measures of family income (square rooted), home ownership, and whether the parents are married that capture other family resources. All standard errors reported here also account for the clustering of multiple respondents within the same household.

# Results

I begin by showing the relative risk of being below expected grade among monoracial respondents across models. @tbl-monoracial reports average marginal effects on the probability of grade retention for each monoracial minority group in comparison to White students. Understanding these differences and how they change across models helps me to understand the potential for how biracial respondents might be positioned.

Model 1 only controls for structural factors such as state of residence, student location, grade, and year. These results provide a baseline estimate of the differences across monoracial groups without controlling for differences in cultural and material resources across groups. The results show that Black, Indigenous, and Latino students all have substantially higher probability of grade retention than White students. Indigenous students have substantially higher risk than all other students, with a probability of grade retention that is 4.1% percentage points higher than White students. Asian students have the lowest probability of grade retention, and their risk is slightly lower than Whites.

The subsequent models control for a variety of additional variables. Model 2 includes controls for whether the student and each of their parents are native-born and speak English well. Model 3 controls for the highest degree received for each parent, and Model 4 controls for family income (square rooted), home ownership, and whether the student's parents are married.

In general Model 2 and Model 3 have substantial effects on the observed inequality across racial groups, while the additional variables from Model 4 have less impact. Controlling for the characteristics related to immigration in Model 2 cuts the gap between White and Latino students in half and increases the gap between Asian and White students. It has no impact on the gap between White and Black or White and Indigenous students because most of these students are native-born with native-born parents.

Controlling for education in Model 3 reduces the gap substantially for Black and Indigenous students and removes the gap entirely for Latino students. In fact, the results of both Models 3 and 4 suggest that, when holding constant cultural and family resources, Latino students are slightly less likely than White students to be held back.

SUMMARIZE THIS?

```{r monoracial-results}
#| results: asis
#| label: tbl-monoracial
#| tbl-cap: Average marginal differences in the probability of being behind grade by monoracial group, relative to a monoracial white student
texreg(map(results_monoracial, convert_marg),
       digits=3,
       caption = "Average marginal differences in the probability of being clearly behind grade by monoracial group, relative to a monoracial white student",
       caption.above=TRUE,
       label="tbl-monoracial",  #TODO: This does not work
       float.pos="t",
       custom.gof.rows=
         list("State fixed effects" = c("Yes","Yes","Yes","Yes"),
              "Location fixed effects" = c("Yes","Yes","Yes","Yes"),
              "Grade fixed effects" = c("Yes","Yes","Yes","Yes"),
              "Year linear effects" = c("Yes","Yes","Yes","Yes"),
              "Year x grade effects"  = c("Yes","Yes","Yes","Yes"),
              "Nativity and language" = c("No","Yes","Yes","Yes"),
              "Parent's education" = c("No","No","Yes","Yes"),
              "Family resources" = c("No","No","No","Yes")))
```

I now turn to the results for biracial respondents. Due to the number of comparisons being made, I split these results into two separate figures. In @fig-black-models, I show the placement of each biracial group involving one black parent in comparison to their two monoracial constituent groups. To allow comparison across models, I compare the results from the baseline model (Model 1 in @tbl-monoracial) and the full model that accounts for all cultural and family resources (Model 4 in @tbl-monoracial).

```{r fig-black-models}
#| label: fig-black-models
#| fig-pos: p
#| fig-cap: Probability of being behind grade for biracial respondents with one black parent, in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects. Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.
#| fig-height: 7.5
coef_table %>%
  filter(model=="Baseline model" | model=="Full model") %>%
  filter(str_detect(mrace, "Black")) %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial 
                             & (model=="Baseline model" | 
                                  model=="Full model") & 
                               str_detect(mrace, "Black")))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average predicted probability of being clearly behind expected grade")
```

I focus first on biracial groups with one Black parent because of a stronger historical expectation that these students will be identified as Black by the "one-drop rule" and as a result will have similar outcomes to Black students. In three cases, this hypothesis can be soundly rejected. For Black/White, Black/Latino, and Black/Asian students, The probability of being clearly below expected grade is substantially lower than that for monoracial black students and the difference is statistically distinguishable from zero. Only in the case of Black/Indigenous students, do I observe outcomes that are indistinguishable from those of Black students. However, in this case,monoracial Black students have lower risk of grade retention than monoracial Indigenous students. Thus, In all four cases, part-Black biracial students have a risk of grade retention that is closer to the lower risk monoracial group.

The results also suggest that each of these biracial groups has a risk of grade retention lower than the halfway point, although this hypothesis cannot be rejected completely for the Black/Asian and Black/Indigenous groups, due to a higher degree of uncertainty for these smaller groups.

Controlling for resource variables has some effect on the placement of Black/White and Black/Latino biracial respondents. In the case of Black/White biracial respondents, Black/White students initially have a risk of grade retention slightly higher than and statistically distinguishable from White students. After controlling for cultural and family resources, however, Black/White respondents have a risk of grade retention that is slightly lower and statistically distinguishable from White students. This change reflects the relatively low level of parental education and income among Black/White students, which is much closer to Black students than White students, as shown on @tbl-descriptive.

An opposite pattern is observed for Black/Latino respondents. Black/Latino respondents initially have a risk of grade retention substantially lower than both Black and Latino students, who have similar risks. However, after controlling for cultural and family resources, Black/Latino students have a risk of grade retention that is indistinguishable from the grade retention risk of Latino students and substantially lower than Black students. The change largely reflects the fact that Black/Latino students have greater cultural resources than Latino students and once this advantage is accounted for, they have a similar risk.

Altogether, the results here strongly suggest that the risk of grade retention faced by these biracial students with one black parent is generally similar to that of their constituent monoracial group with the lower risk of grade retention. This finding is inconsistent both with the expectation of the one-drop rule and the "in-between" hypothesis.

@fig-other-models shows the results for the remaining biracial groups. In general, the results here are consistent with those for part-Black biracial respondents, with some important caveats discussed below. The point estimates indicate that the risk of each biracial group is either quite close to the monoracial group with the lower risk or in-between this group's risk and the halfway point.

```{r fig-other-models}
#| label: fig-other-models
#| fig-pos: p
#| fig-cap: Probability of being behind grade for non-black biracial respondents, in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects. Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.
#| fig-height: 7.5
coef_table %>%
  filter(model=="Baseline model" | model=="Full model") %>%
  filter(!str_detect(mrace, "Black")) %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial 
                             & (model=="Baseline model" | 
                                  model=="Full model") & 
                               !str_detect(mrace, "Black")))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average predicted probability of being clearly behind expected grade")
```

For all groups, the full models generally indicate that each biracial group has a risk of grade retention similar to the risk of the lower monoracial constituent group or has risk higher than this group but less than the halfway point. There are two exceptions to this general pattern. First, for the Indigenous/Latino group, I cannot rule out the possibility that this group has risk roughly halfway between the two constituent groups, although the point estimate favors a result closer to the lower Latino group. Second, Latino/Asian students are the only case where the risk for the biracial group is higher than either monoracial group. However, the confidence bands suggest that the risk for the Latino/Asian group cannot be differentiated from the rates for Latinos, the higher of the two monoracial constituent groups. This result marks the one substantial deviation from the general pattern. It only emerges in the models that control for cultural and family resources because Latino/Asian students come from higher income and more educated households than Latino students, on average.

Aside from the effect on Latino/Asian students, controlling for cultural and family resources has moderate effects on the placement of the remaining biracial groups. The only case with a substantial shift is that for White/Asian students who initially appear to have lower risk than either White or Asian students. After controlling for these resources, the risks of White/Asian students are indistinguishable from the lower risk of Asian students. This is due to the fact that White/Asian students come from the most advantaged households of any group in the dataset.

Any results for part-Latino students are complicated by the substantial change in the monoracial Latino risk across models. For example, while Latinos have substantially higher risk of grade retention than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model. Nonetheless, even in this case, we see that White/Latinos have risk closer to those of the lower Latino group, but distinguishable from both groups.

# Conclusions

-   Big claim: The results are consistent with a new "regime of hyperdescent" operating across all racial groups. This contributes to a muddying of racial distinctiveness via hybridity, while at the same time isolating monoracial members of the most stigmatized racial groups, consistent with arguments like Bonilla-Silva's Latin Americanization.

# Acknowledgments {.appendix}

I am grateful to...

# References {.appendix}
