---
title: Grade Retention and Mixed Race
author: 
  - Aaron Gullickson, *University of Oregon*
abstract: This is a test abstract
thanks: Thanks to people and stuff
format:
  pdf:
    documentclass: scrartcl
    papersize: letter
    fig-height: 4
    fig-pos: t
    keep-tex: false
bibliography: ../project.bib
execute: 
  eval: true
  echo: false
  warning: false
  error: false
editor: source
---

```{r setup}
#| include: false
library(here)
source(here("analysis","check_packages.R"))
source(here("analysis","useful_functions.R"))
```

```{r load-data}
load(here("analysis","output","acs.RData"))
load(here("analysis","output","coef_table.RData"))
load(here("analysis","output","monoracial_results.RData"))
# change names of models for display
coef_table <- coef_table %>%
  mutate(model=factor(model,
                      levels=c("crude","fixed effects","cultural resources",
                               "parent resources"),
                      labels=c("Raw","Baseline model","Cultural model",
                               "Full model")))
```

# Introduction

* Research on mixed-race individuals in the US has exploded over the last two decades, driven by both the demographic phenomenon of a biracial baby boom and changes to statistical reporting that allow multiple race reporting. 
* Although, multiraciality in the US is a phenomenon that is in fact older than the country itself (CITATIONS), much of the research has focused particularly on the experiences of first-generation "biracial" individuals whose parents belong to different single race groups. The reasons for this are:
  * These individuals are coming of age in a post-Civil Rights era in which historical norms of racial identification may be dissolving and/or shifting.
  * The demographic weight of the biracial baby boom suggests that these individuals will play an outsized role in the future of race in the US.
* Research has focused on a variety of areas with regard to this group. In particular, studies of how such individuals identify and are classified by others has been extremely active. Significantly less work has been done on the life chances of first-generation biracials relative to their monoracial peers.
* Such work is critically important because it helps us understand how contemporary biracials are sorted into a pre-existing system of structural racial inequality. 
* Researchers are well aware of the importance of this question, the relative lack of work on the topic reflects two important limitations in our data sources.
  * First, simply using self-identification in large scale sources like the Census, ACS, or CPS data is generally seen as problematic. We know that many biracial individuals will not identify as multiracial and some individuals who identify as multiracial will not be first-generation biracials. Thus, any observed differences across groups may be the result of selection biases in who decides to identify as multiracial. 
  * Second, other data sources that allow identification of biracial respondents by parental race often contain samples of biracial respondents that are too small to facilitate analysis. This is particularly true when researchers want to consider outcomes separately by the specific parental racial groups involved, as they should.
* One way that researchers have overcome this limitation in prior work is by the identification of biracial children living with both biological parents in large-scale Census and ACS data. This approach has been primarily used to look at how such children are racially classified (citations), but has also been used to look at residential settlement patterns (citations). Its more difficult to look at traditional educational and labor market outcomes, however, precisely because the biracial respondents identified will be minors and thus will not have accumulated much in the way of life outcomes, for good or bad. 
* One important outcome is available, however, for children. Grade retention (e.g. being held back in school). We know that grade retention is relatively common in the US and that the likelihood of grade retention is not equally distributed for children but is rather related to socioeconomic status, family structure, and race (citations). We also know that grade retention is related to later negative educational and labor market outcomes (citations). Thus, grade retention provides one of the earliest measures of the way systems of stratification sort individuals into a system of structural inequality. 
* In this study, I analyze the grade retention outcomes of biracial children relative to their monoracial peers using ACS data from 2010-2019. The results provide a window into the placement of biracial children within an existing system of racial inequality in our educational systems. ==TAKING TOO LONG TO GET HERE, MIGHT WANT TO RE-ORDER SOME THINGS==

# Background

* The biracial baby boom - why we should care about outcomes for this particular group. 

* What should we expect about the outcomes for biracials relative to their monoracial peer groups?
  * In each case, our interest is in the outcomes for the biracial group relative to the two constituent groups that their parents identify with. So for example, how to the outcomes of a Black/White individual compare to monoracial Black and monoracial White individuals?
  * The most direct pathway for understanding these outcomes is via how such individuals are classified by those in positions of power and control, such as teachers, administrators, employers, etc. To the extent that biracial individuals are classified as more like one of their monoracial constituent groups, they will receive the same treatment by individuals and either reap similar benefits or suffer similar discrimination and penalties.
  * Specifically, I consider two long-standing norms of *hypodescent* and *hyperdescent* in understanding possible placement of biracial individuals (cite F James Davis). Hypodescent indicates a norm in which individuals of mixed racial ancestry are classified as members of the lower status group within a racial hierarchy. In the US, hypodescent is canonically tied to the "one-drop rule", a norm in which those of part-Black ancestry were simply viewed as Black, irrespective of other ancestry (CITATIONS). More generally, if hypodescent is the common practice, we would expect biracial individual's outcomes to be closest those of the monoracial constituent group with the lower outcomes. 
  * Hyperdescent refers to the opposite situation in which a mixed race individual is classified as a member of the higher status group. Because hyperdescent blends individuals into the often unmarked dominant group, it is often more difficult to see. Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming membership with another group (most notably Whites). In the case of outcomes, a pattern of hyperdescent would be indicated if biracial individual's outcomes are closest to those of the monoracial constituent group with the higher outcomes. 
  * Discuss work on the contemporary strength of these norms? 
  * The expectations of hypodescent/hyperdescent are that the biracial respondent will have outcomes close to one of the monoracial constituent groups. Another possibility is that biracial respondents will have outcomes intermediate between their monoracial constituent groups. There could be several reasons for this:
    * The racial ambiguity of biracial respondents may protect them from the worst stigmatization and discrimination in social settings, but by the same coin it may allow them to benefit less from membership in advantaged groups. Thus, a greater recognition of their "blended" ancestry may place them in a halfway position.
    * Even in cases where hypodescent and hyperdescent are still norms, physical differentation such as skin tone may provide advantages and disadvantages among members who are classified in the same way. We know that skin tone makes a difference even for monoracial Blacks, so the one-drop rule while providing a single classification does not create uniform experiences for all Blacks. (==CITE MONKS WORK HERE I THINK==).
    * Successful outcomes within social systems depend on a variety of non-material, symbolic resources (e.g. social and cultural capital) that we rarely completely capture in our measurements. Some of the differences in outcomes across racial groups may reflect differences in those resources (CITATIONS?). Given that biracial children have parents belonging to two different races, they may possess such resources that are roughly half-way between the two constituent groups on average, and thus will place them roughly in between the groups. In effect, the in-between outcomes for biracial children may loosely reflect the "black box" package of differences across racial groups that a racial dummy variables stands in for in our research practices.
    * Two other possibilities don't suggest the actual placement, but rather may modify it in the face of other prevalent patterns. Hybridity itself may be a boon or a burden on biracial individuals that affects their outcomes in, respectively, positive or negative ways. 
      * On the positive side of the ledger, hybridity may enabled a larger cultural "toolkit" that allows mixed race individuals to more easily navigate a variety of social situations. On the negative side of the ledger, the longstanding "marginal man" hypothesis suggests that the dissonance of living in two different worlds may lead to negative mental health outcomes for mixed race individuals, which could spill over into other types of outcomes.

* In all of this, its crucially important to understand how other resources affect our understanding of the placement of biracial individuals. Biracial individuals are the products of interracial unions and individuals are not randomly selected into such unions. This can lead to underlying differences in the socioeconomic resources that biracial individuals possess relative to their monoracial peers. However, this selection can be quite complex. (DISCUSS CHENG AND LIVELY HERE?)
  * Status exchange suggests that members of lower status racial groups will be positively selected into interracial unions, while higher status racial groups will be negatively selected. How this affects the relative resources of biracial individuals will depend then in some part on the relative strength of these phenomena, the gender balance in a given form of interracial unions, and how the gender of parents affects outcomes. Status exchange may also not be present in all forms of interracial unions.

* Review previous studies on outcomes
  * Campbell
  * Herman
  * Holloway/Ellis stuff
  * Udry? - also related stuff by Cheng? How does Cheng and Powell fit in here?
  * Do some more bibliographic work, particularly forward searching to see if I am missing something big here.
  * Limitations of this work
    * Small sample sizes lead to inconsistent estimates in most of the survey data.
    * In many cases, impossible to understand outcomes for part-Latino biracials. 
  
* Why grade retention as an outcome?
  * Grade retention is a relatively common practice. Discuss overall trends in grade retention and connection to educational reforms.
  * Grade retention has negative effect on later outcomes, even controlling for cognitive differences (CITE WORK HERE).
  * Clear inequalities by race in practices of grade retention as well as by SES (CITE WORK).
  * In effect, grade retention may be for many individuals the first encounter with the wider stratification system that sorts individuals and helps to perpetuate inequality.
  
* What I do  and how it improves our understanding.

# Data and Methods

Data for this analysis come from the American Community Survey (ACS), an annual 1-in-100 survey of the US population, produced by the Census Bureau. To increase sample size for smaller populations of biracial respondents, I pool ACS data for a full decade from 2010-2019. While pooling data helps to increase sample size, it may generate biases if the likelihood of grade retention is changing over time in complex ways. I address this issue in the modeling techniques discussed below. All data were extracted from the Minnesota Population Center's Integrated Public Use Microdata Series [@ruggles_ipums_2020].

Ideally, I would restrict the sample only to children living in a household with two biological parents. However, the ACS only records the direct family relationship between the head of household andother members of the family. Since 2008, the ACS also distinguishes between biological children, adopted children, and stepchildren. Respondents can therefore be identified as biological children of another member of the household only if that member is the head of household or the biological child of the head of household. Determining whether the partner/spouse of that biological parent is the other biological parent is more difficult.

Prior work has used a variety of additional restrictions to limit the analysis to children who are more likely to be the biological children of both parents [@saenz_persistence_1995; @xie_racial_1997]. Some of those procedures rely upon an assumption that most parents would be married that was more tenable in historical data than contemporary data. Therefore, I do not explicitly restrict by marital status, I do restrict cases in two other ways that are consistent with prior work. First, I restrict the sample to cases where both parents were of a reasonable age at the birth of the child (aged 15-44 for mothers and aged 15-60 for fathers). Second, I restrict the sample to only those cases where the reported race and Hispanicity of the child is consistent with at least one of the parents.

The final analytical sample consists of children aged 5-20 who are currently enrolled in the K-12 school system and live in a household with two parents who are both likely to be biological parents. This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren, limiting it exclusively to households with two parents present and over-representing the proportion of cases in which those parents are married. The result of this bias is well-known. Children living in two-parent nuclear households have greater educational success than other children as a result of material, social, and cultural resources (CITATIONS). Therefore, this sample restriction will upwardly bias estimates of grade retention. However, this bias is less problematic for the goal of this study, which is to estimate differences between groups. The sample restriction effectively eliminates differences in family structure across racial groups which may account for some observed differences in grade retention. However, the remaining differences observed here are likely to hold for the full population in the absence of strong racial differences in the effect of family structure on grade retention.

The race of children is calculated by a cross-tabulation of parent's races. A full cross-tabulation of the race/Hispanicity of each parent would produce far too many combined categories to be analytically useful. Therefore, I collapse each parent's race into the categories of White, Black, Asian, Latino, and Indigenous. These categories form the "ethnoracial pentagon" that is commonly used in popular practice and government tabulation to identify race in the US [@hollinger_postethnic_1995]. Respondents are identified as Latino base on their response to the Hispanicity question, regardless of their response to the race question. Respondents are identified as Indigenous if they responded as either American Indian/Alaska Native or as Pacific Islander. Because the goal of my analysis is to examine specifically the outcomes of "first-generation" biracial children, I exclude cases where at least one parent identified with multiple races.

The cross-tabulation of parent's races leads to ten distinct biracial categories. The total sample size for each of these biracial categories is shown in @tbl-descriptive, along with several descriptive statistics. The sample sizes are quite large for many of the groups, suggesting substantial statistical power in determining the relative outcomes for these groups. Even among the smallest group of Indigenous/Asian, the sample size is substantially greater than one would see in typical survey data.

```{r}
#| label: tbl-descriptive
#| tbl-cap: Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.
tbl_summary <- acs %>%
  group_by(race) %>%
  summarize(n=n(), 
            held_back=percent(mean(below_exp_grade), accuracy=0.01),
            mean_age=round(mean(age), 1),
            f_born=percent(mean(foreign_born),  accuracy=0.1),
            par_fb=percent(mean(foreign_born_father | 
                                  foreign_born_mother), accuracy=0.1),
            fam_inc=dollar(round(mean(family_income), 0)),
            own_home=percent(mean(own_home), accuracy=0.1),
            par_ba=percent(mean(degree_father=="BA" | 
                                  degree_father=="G" | 
                                  degree_mother=="BA" | 
                                  degree_mother=="G"), accuracy=0.1)
  ) %>%
  arrange(desc(n)) %>%
  mutate(n=comma(n))

index_biracial <- which(str_detect(tbl_summary$race, "/"))

tbl_summary %>%
  kbl(booktabs = TRUE,
      linesep = NULL,
      align = c("l",rep("r", ncol(tbl_summary)-1)),
      col.names=c("race",
                  "n",
                  "clearly below expected grade", 
                  "mean age",
                  "foreign-born",
                  "either parent foreign-born",
                  "mean family income",
                  "own home",
                  "either parent, college degree"
      )) %>%
  kable_styling(font_size=8) %>%
  column_spec(3:10, width = "1.2cm") %>%
  row_spec(index_biracial, background = "#D3D3D3") 
#%>% landscape()
```

To measure grade retention, I compare the age of respondents to their current grade in school. Because the ACS lacks data on the exact birthday of respondents, it is not possible to exactly identify students who are below the expected grade level. A student of any given age in years can always be expected in two different grades. For example, a first grader is expected to be either six or seven years of age, while a second grader is expected to be either seven or eight years of age. Therefore, with information only on current grade and age in years, it is impossible to determine whether a seven year old in first grade is behind the expected grade level.

Instead of trying to identify exact grade retention, I instead identify cases in which students are clearly below expected grade. A student is clearly behind expected grade when the age of the student is greater than either of the possible ages for a student of that grade (e.g. a first grader who is eight years old). This measure will underestimate the total proportion of students that have been held back. However, the goal of this study is to examine differences across groups rather than total levels. The only factor determining this bias is whether the birthday of the student takes place earlier or later in the year, and is unlikely to vary substantially across groups.[^1]

[^1]: Biracial respondents are slightly younger than monoracial respondents in the sample as a whole because interracial unions are becoming more common over time HOW STRONG COULD THE BIAS BE HERE AND WHAT CAN I DO ABOUT IT?

==ACKNOWLEDGE POTENTIAL ISSUE WITH DROPOUT STUDENTS==

@tbl-descriptive shows the percent of each monoracial and biracial group that is clearly below the expected grade. My goal is to understand the degree of grade retention for each biracial group relative to its constituent monoracial groups (e.g. a Black/White student compared to White and Black students). Throughout this article, I use a visual approach to illustrate this placement.

As an example, I display the percent of students clearly behind grade for White, Black, and Black/White students in @fig-example. The point estimate and confidence interval for the Black/White group are shown in red, while the two monoracial groups are shown via grey dots and confidence bands. For all three cases, I use 83.4% confidence intervals, because overlap on such intervals indicates that the observed difference between the two measures in the sample is not statistically distinguishable from zero in the population at $p<.05$ (CITE).[^2] In this case, the lack of overlap between any of the confidence bands indicates that the Black/White probability of being clearly below expected grade is statistically distinguishable from both monoracial probabilities.

[^2]: The correspondence to $p<.05$ is only approximate due to differences in the two standard errors between estimates. However, this approximation will be extremely close in almost all cases.

```{r example-plot}
#| label: fig-example
#| fig-cap: Probability of being clearly below expected grade for biracial black/white respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
g <-  coef_table %>%
  filter(mrace=="Black/White" & model=="Raw") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial & 
                               mrace=="Black/White" &
                               model=="Raw"))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade")

g+annotate("text", 1, 0.035, hjust="left",
             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
  annotate("text", 2, 0.045, hjust="left",
             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
  annotate("text", 2.5, 0.036, hjust="left",
             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
```

Additionally, I include a measure of the "halfway" point between the two racial groups.[^3] This halfway point allows me to determine where the biracial group falls relative to the expectation of being halfway between the two constituent monoracial groups. In this case, Black/White students are much closer to White students in their risk of grade retention, and have probabilities much lower than both Black students and the halfway expectation.

[^3]: The standard error of this halfway estimate used to construct the confidence interval is given by $\sqrt{(s_1^2+s_2^2)/2}$ where $s_1$ and $s_2$ are the standard errors of the estimates for the two monoracial groups.

@fig-example shows the raw differences between the three racial groups of interest. However, in practice, I want to estimate differences across groups while holding constant a variety of potentially confounding variables. To do this, I estimate a set of logit models that predict the likelihood of a student being clearly behind expected grade by race and a variety of other variables. I then construct figures similar to @fig-example by calculating from the model the average predicted probabilities (APP) for each racial group. APPs estimate the average probability of the outcome by some covariate across all cases, while holding constant all other variables. APPs are akin to average marginal effects (AME) and estimated in the same way via the delta method. Whereas AMEs estimate differences or slopes, APPs estimate the level for a given category.[^4]

[^4]: Output from the full models used to construct these APPs is available in the supplementary materials provided with this article.

I begin with a baseline model that adjusts for a variety of nuisance characteristics that may vary across groups and thus need to be controlled in all models. First, I include fixed effects for state of residence because states dictate educational policy and thus can differ substantially in the likelihood of grade retention. I cannot identify specific school districts, but I also include dummy variables indicating whether the student lived in a central city, suburban, or rural area.[^5]

[^5]: The location cannot be determined for all students and I therefore also include a fourth category of unknown in all models.

The probability of being clearly below expected grade also increases with the student's current grade, so I include fixed effects for the current grade of the student. Additionally, @fig-year-grade shows that the percent of students clearly behind grade level has declined over time, but this decline has been much more substantial at higher grade levels. For the elementary grades, there is no evidence of a decline at all. To account for this grade-specific decline in the models I include an interaction between a linear year term and current grade.[^6]

[^6]: Sensitivity analysis showed that this functional form was preferred by the Bayesian information criterion to models with no interaction and a model with interaction terms that treated year as a categorical variable. I also considered similar interactions between state and year, but none of the models with these interaction terms were preferred.

```{r}
#| label: fig-year-grade
#| fig-cap: Trends over time and grade in the percent of students clearly behind grade level. Lines are fit to each set of points by grade via lowess smoothing.
  
acs %>%
  group_by(current_grade, year) %>%
  summarize(below_grade=mean(below_exp_grade)) %>%
  ggplot(aes(x=year, y=below_grade, group=current_grade, color=current_grade))+
  #geom_line()+
  geom_smooth(se=FALSE)+
  geom_point()+
  scale_x_continuous(breaks=c(2010,2014,2018))+
  scale_y_continuous(labels=scales::percent)+
  labs(x="year", y="percent clearly behind grade level for age",
       color="current grade")+
  theme_bw()
```

The baseline model also includes the race of the respondent. I then add additional terms measuring material and cultural resources that may account for racial differences in a set of subsequent models. First, I include measures of nativity and English proficiency for both the respondent and each of their parents. Second, I include a categorical measure of highest degree earned for each parent. Finally, I include measures of family income (square rooted), home ownership, and whether the parents are married that capture other family resources. All standard errors reported here also account for the clustering of multiple respondents within the same household.

# Results

I begin by showing the relative risk of being below expected grade among monoracial respondents across models. @tbl-monoracial reports average marginal effects on the probability of grade retention for each monoracial minority group in comparison to White students. Understanding these differences and how they change across models helps me to understand the potential for how biracial respondents might be positioned.

Model 1 only controls for structural factors such as state of residence, student location, grade, and year. These results provide a baseline estimate of the differences across monoracial groups without controlling for differences in cultural and material resources across groups. The results show that Black, Indigenous, and Latino students all have substantially higher probability of grade retention than White students. Indigenous students have substantially higher risk than all other students, with a probability of grade retention that is 4.1% percentage points higher than White students. Asian students have the lowest probability of grade retention, and their risk is slightly lower than Whites.

The subsequent models control for a variety of additional variables. Model 2 includes controls for whether the student and each of their parents are native-born and speak English well. Model 3 controls for the highest degree received for each parent, and Model 4 controls for family income (square rooted), home ownership, and whether the student's parents are married.

In general Model 2 and Model 3 have substantial effects on the observed inequality across racial groups, while the additional variables from Model 4 have less impact. Controlling for the characteristics related to immigration in Model 2 cuts the gap between White and Latino students in half and increases the gap between Asian and White students. It has no impact on the gap between White and Black or White and Indigenous students because most of these students are native-born with native-born parents.

Controlling for education in Model 3 reduces the gap substantially for Black and Indigenous students and removes the gap entirely for Latino students. In fact, the results of both Models 3 and 4 suggest that, when holding constant cultural and family resources, Latino students are slightly less likely than White students to be held back.

SUMMARIZE THIS?

```{r monoracial-results}
#| results: asis
#| label: tbl-monoracial
#| tbl-cap: Average marginal differences in the probability of being behind grade by monoracial group, relative to a monoracial white student
texreg(map(results_monoracial, convert_marg),
       digits=3,
       caption = "Average marginal differences in the probability of being clearly behind grade by monoracial group, relative to a monoracial white student",
       caption.above=TRUE,
       label="tbl-monoracial",  #TODO: This does not work
       float.pos="t",
       custom.gof.rows=
         list("State fixed effects" = c("Yes","Yes","Yes","Yes"),
              "Location fixed effects" = c("Yes","Yes","Yes","Yes"),
              "Grade fixed effects" = c("Yes","Yes","Yes","Yes"),
              "Year linear effects" = c("Yes","Yes","Yes","Yes"),
              "Year x grade effects"  = c("Yes","Yes","Yes","Yes"),
              "Nativity and language" = c("No","Yes","Yes","Yes"),
              "Parent's education" = c("No","No","Yes","Yes"),
              "Family resources" = c("No","No","No","Yes")))
```

I now turn to the results for biracial respondents. Due to the number of comparisons being made, I split these results into two separate figures. In @fig-black-models, I show the placement of each biracial group involving one black parent in comparison to their two monoracial constituent groups. To allow comparison across models, I compare the results from the baseline model (Model 1 in @tbl-monoracial) and the full model that accounts for all cultural and family resources (Model 4 in @tbl-monoracial).

```{r fig-black-models}
#| label: fig-black-models
#| fig-pos: p
#| fig-cap: Probability of being behind grade for biracial respondents with one black parent, in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects. Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.
#| fig-height: 7.5
coef_table %>%
  filter(model=="Baseline model" | model=="Full model") %>%
  filter(str_detect(mrace, "Black")) %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial 
                             & (model=="Baseline model" | 
                                  model=="Full model") & 
                               str_detect(mrace, "Black")))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average predicted probability of being clearly behind expected grade")
```

I focus first on biracial groups with one Black parent because of a stronger historical expectation that these students will be identified as Black by the "one-drop rule" and as a result will have similar outcomes to Black students. In three cases, this hypothesis can be soundly rejected. For Black/White, Black/Latino, and Black/Asian students, The probability of being clearly below expected grade is substantially lower than that for monoracial black students and the difference is statistically distinguishable from zero. Only in the case of Black/Indigenous students, do I observe outcomes that are indistinguishable from those of Black students. However, in this case,monoracial Black students have lower risk of grade retention than monoracial Indigenous students. Thus, In all four cases, part-Black biracial students have a risk of grade retention that is closer to the lower risk monoracial group.

The results also suggest that each of these biracial groups has a risk of grade retention lower than the halfway point, although this hypothesis cannot be rejected completely for the Black/Asian and Black/Indigenous groups, due to a higher degree of uncertainty for these smaller groups.

Controlling for resource variables has some effect on the placement of Black/White and Black/Latino biracial respondents. In the case of Black/White biracial respondents, Black/White students initially have a risk of grade retention slightly higher than and statistically distinguishable from White students. After controlling for cultural and family resources, however, Black/White respondents have a risk of grade retention that is slightly lower and statistically distinguishable from White students. This change reflects the relatively low level of parental education and income among Black/White students, which is much closer to Black students than White students, as shown on @tbl-descriptive.

An opposite pattern is observed for Black/Latino respondents. Black/Latino respondents initially have a risk of grade retention substantially lower than both Black and Latino students, who have similar risks. However, after controlling for cultural and family resources, Black/Latino students have a risk of grade retention that is indistinguishable from the grade retention risk of Latino students and substantially lower than Black students. The change largely reflects the fact that Black/Latino students have greater cultural resources than Latino students and once this advantage is accounted for, they have a similar risk.

Altogether, the results here strongly suggest that the risk of grade retention faced by these biracial students with one black parent is generally similar to that of their constituent monoracial group with the lower risk of grade retention. This finding is inconsistent both with the expectation of the one-drop rule and the "in-between" hypothesis.

@fig-other-models shows the results for the remaining biracial groups. In general, the results here are consistent with those for part-Black biracial respondents, with some important caveats discussed below. The point estimates indicate that the risk of each biracial group is either quite close to the monoracial group with the lower risk or in-between this group's risk and the halfway point.

```{r fig-other-models}
#| label: fig-other-models
#| fig-pos: p
#| fig-cap: Probability of being behind grade for non-black biracial respondents, in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects. Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.
#| fig-height: 7.5
coef_table %>%
  filter(model=="Baseline model" | model=="Full model") %>%
  filter(!str_detect(mrace, "Black")) %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial 
                             & (model=="Baseline model" | 
                                  model=="Full model") & 
                               !str_detect(mrace, "Black")))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average predicted probability of being clearly behind expected grade")
```

For all groups, the full models generally indicate that each biracial group has a risk of grade retention similar to the risk of the lower monoracial constituent group or has risk higher than this group but less than the halfway point. There are two exceptions to this general pattern. First, for the Indigenous/Latino group, I cannot rule out the possibility that this group has risk roughly halfway between the two constituent groups, although the point estimate favors a result closer to the lower Latino group. Second, Latino/Asian students are the only case where the risk for the biracial group is higher than either monoracial group. However, the confidence bands suggest that the risk for the Latino/Asian group cannot be differentiated from the rates for Latinos, the higher of the two monoracial constituent groups. This result marks the one substantial deviation from the general pattern. It only emerges in the models that control for cultural and family resources because Latino/Asian students come from higher income and more educated households than Latino students, on average.

Aside from the effect on Latino/Asian students, controlling for cultural and family resources has moderate effects on the placement of the remaining biracial groups. The only case with a substantial shift is that for White/Asian students who initially appear to have lower risk than either White or Asian students. After controlling for these resources, the risks of White/Asian students are indistinguishable from the lower risk of Asian students. This is due to the fact that White/Asian students come from the most advantaged households of any group in the dataset.

Any results for part-Latino students are complicated by the substantial change in the monoracial Latino risk across models. For example, while Latinos have substantially higher risk of grade retention than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model. Nonetheless, even in this case, we see that White/Latinos have risk closer to those of the lower Latino group, but distinguishable from both groups.

## Conclusions

* Big claim: The results are consistent with a new "regime of hyperdescent" operating across all racial groups. This contributes to a muddying of racial distinctiveness via hybridity, while at the same time isolating monoracial members of the most stigmatized racial groups, consistent with arguments like Bonilla-Silva's Latin Americanization. 

## References
